---
output:
  html_document: 
  html_notebook:
    fig_caption: yes
    
---
# Reproducible Research: Peer Assessment 1
---
title: 'Reproducible Research: Peer Assessment 1'
author: "Mahendra S Payaal"
date: "4/2/2020"


     

---

##  Setting Global Options

```{r savingplots}
knitr::opts_chunk$set(fig.path = "figure/")
library(markdown)
```

##  Loading and preprocessing the data
```{r loadingData}
activity <- read.csv("C:/Users/mahendra.payaal/Documents/coursera/data/activity.csv",stringsAsFactors = FALSE)
activity  <- transform( activity, date = as.Date(date,"%Y-%m-%d"))

```

##  What is mean total number of steps taken per day?

###    Total number of steps taken everyday

```{r totalDailysteps}
steps_daily_Total <- as.data.frame(with(activity,tapply(steps,date,sum,na.rm=TRUE)))
rownames(steps_daily_Total) <-c(1:61)
colnames(steps_daily_Total) <- "Daily_Steps_Sum"
steps_daily_Total
```

###    Histogram of the total number of steps taken each day

```{r histogram_of_total_daily_steps}
hist(steps_daily_Total$Daily_Steps_Sum,main ="Histogram of Total Daily Steps over Two Months", xlab = "Daily Steps")
abline(v=mean(steps_daily_Total$Daily_Steps_Sum,na.rm = TRUE),lwd =2,col="blue")
text(8000,15,"Mean",col="blue")

abline(v=median(steps_daily_Total$Daily_Steps_Sum,na.rm = TRUE),lwd =2,col="magenta")
text(12000,25,"Median",col="magenta")
```

###    Mean total number of steps taken per day
```{r}
mean_daily <- mean(steps_daily_Total$Daily_Steps_Sum,na.rm = TRUE)
```
      The mean of total daily steps is `r mean_daily`

###    Median of total daily steps 
```{r}
median_daily <- median(steps_daily_Total$Daily_Steps_Sum)
```
       The median of total daily steps is `r median_daily`

##  What is the average daily activity pattern?

###    Time series plot of the average number of steps taken

```{r plot_average_steps_by_interval}
steps_interval_Average <- as.data.frame(with(activity,tapply(steps,interval,mean,na.rm=TRUE)))
names(steps_interval_Average) <- "Average_steps"
plot(as.numeric(rownames(steps_interval_Average)),steps_interval_Average$Average_steps,type ="l",xlab="24 Hour Interval of 5 Min each",ylab="Number of Steps")
title("Time Plot of Average Steps in Every 5 Minutes Interval")
```

###    The 5-minute interval that, on average, contains the maximum number of steps

```{r}
max_interval <- which.max(steps_interval_Average[,1])
names_interval <- rownames(steps_interval_Average)
max_interval<-names_interval[104]
max_int <- sub("(\\d+)(\\d{2})","\\1:\\2",max_interval)
```
      The average maximun  steps are in 5 minute interval from `r max_int`

##  Imputing missing values

###    Total number of missing values 

```{r}
total_NA <- sum(is.na(activity$steps))
```

       The total NAs in data frame  are `r total_NA`

###    Code to describe and show a strategy for imputing missing data and creating a new dataset

- 'NA' values are replaced by the average values for that time interval. The steps are given below
- A vector"avg_steps" is created which has average value of  all intervals
- This vector is replicated 61 times to become equal to rows of "activity"  dataframe
- A new  data set "activity_new" is created
- Through a 'For' loop 'NA' values are replaced by average values of that interval

```{r}
avg_steps <- steps_interval_Average[,1]
avg_steps <- c(rep(avg_steps,times =61))
activity_new <- activity
for (i in 1:nrow(activity_new)) {
      if(is.na(activity_new$steps[i])){
                activity_new$steps[i] <- avg_steps[i]
                 }
}
steps_daily_Total_new <- as.data.frame(with(activity_new,tapply(steps,date,sum,na.rm=TRUE)))
rownames(steps_daily_Total_new) <-c(1:61)
colnames(steps_daily_Total_new) <- "Daily_Steps_Sum"
```

###    Histogram of the total number of steps taken each day after missing values are imputed

```{r histogram_total_daily_steps_imputed}
hist(steps_daily_Total_new$Daily_Steps_Sum,main ="Histogram of Total Daily Steps after Imputing", xlab = "Daily Steps")
abline(v=mean(steps_daily_Total_new$Daily_Steps_Sum,na.rm = TRUE),lwd =2,col="blue")
text(9000,15,"Mean",col="blue")
text(12225,25,"Median",col="magenta")
abline(v=median(steps_daily_Total_new$Daily_Steps_Sum,na.rm = TRUE),lwd =2,col="magenta")
```

###    Mean of  total number of steps taken per day after imputing

```{r}
mean_daily_imputed <-mean(steps_daily_Total_new$Daily_Steps_Sum,na.rm = TRUE)
```

       The mean of total daily steps after imputing is `r mean_daily_imputed`

###    Median of  total number of steps taken per day after imputing

```{r}
median_daily_imputed <- median(steps_daily_Total_new$Daily_Steps_Sum)
```
 
      The median of total daily steps after imputing is `r median_daily_imputed`
      
      The mean and median earlier were 9354.23 and 10395. After imputing the mean and median has increased and become equal, and the new value is 10766.19 

##  Are there differences in activity patterns between weekdays and weekends?

###    Creating a new factor variable based on weekday and weekend

- The 'date' variable of "activity_new" dataframe has been identified as "weekday" and "weekend"
- The date variable is replaced with "weekday" or "weekend"
- Two subsets of new dataset -"activity_new", are created based on "weekday" and "weekend"
- Average is taken by "tapply" across all intervals for the days for two subsets

```{r}
wday <- weekdays(activity_new$date)
for( i in 1:17568) {
    if(wday[i] =="Saturday"|wday[i]=="Sunday"){
         wday[i] <-"weekend"
    }
  else {wday[i] <-"weekday" }
  
}
activity_new$date <- wday
weekday_steps <- subset(activity_new,date=="weekday")
weekend_steps <- subset(activity_new,date=="weekend")
steps__weekday <- as.data.frame(with(weekday_steps,tapply(steps,interval,mean,na.rm=TRUE)))
steps__weekend <- as.data.frame(with(weekend_steps,tapply(steps,interval,mean,na.rm=TRUE)))
names(steps__weekday) <- "Average_steps"
names(steps__weekend) <- "Average_steps"
```

###   Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r weekday versus weekend }
par(mfrow=c(1,2),mar=c(4,4,3,1))
plot(as.numeric(rownames(steps__weekday)),steps__weekday$Average_steps,type ="l",xlab="24 Hour Interval of 5 Min each",ylab="Average Steps")
title("Time Plot for Weekdays")

plot(as.numeric(rownames(steps__weekend)),steps__weekend$Average_steps,type ="l",xlab="24 Hour Interval of 5 Min each",ylab="Average Steps",xlim =c(0000,2359))
title("Time Plot  for Weekend")
```

